---
title: "Getting started: from data upload to plot export"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: from data upload to plot export}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This guide walks through the current ADaMViz SubmissionSync proof of concept:
upload a synthetic dataset, confirm a longitudinal boxplot, inspect the result,
and download the plot and R script.

The example uses synthetic ALT measurements. The supported plot pattern is a
longitudinal numeric BDS boxplot. Other plot types and broader table and listing
workflows are outside this walkthrough.

## Before you start

Use the supplied synthetic file for this exercise. The POC permits public,
synthetic, or properly de-identified data; confidential patient-level data is
prohibited. Upload checks establish format compatibility, not de-identification
or permission to use a dataset.

The application is designed for validated clinical submission workflows.
It is not FDA validated or FDA approved. This walkthrough demonstrates the POC;
it does not establish statistical or intended-use validation.

From the repository root, open the RStudio Console. If the package is already
loaded after source changes, restart R before installing it:

```{r launch, eval=FALSE}
devtools::install()
ADaMViz.SubmissionSync::run_app()
```

If the project environment has not been restored, follow the repository's
environment setup before installing. The application launches interactively;
the vignette does not start a Shiny session while rendering.

Locate the example file in the installed package:

```{r sample-path, eval=FALSE}
sample_file <- system.file(
  "extdata", "synthetic", "adlb-standard.csv",
  package = "ADaMViz.SubmissionSync",
  mustWork = TRUE
)
sample_file
```

In a source checkout, the same file is at
`inst/extdata/synthetic/adlb-standard.csv`. It contains 18 records, one parameter
(`ALT`), two treatment groups (`Placebo` and `Active`), and three visits
(`Baseline`, `Week 4`, and `Week 8`).

## 1. Upload the data

On the **Data** step, choose **Choose file** under **Upload a dataset** and
select `adlb-standard.csv`. Wait for **File ready**, then select **Continue**.

The application accepts one CSV or Excel (`.xls` or `.xlsx`) file at a time.
For Excel, it reads the first worksheet. The data must be in long format, with
subject, parameter, and visit information stored in columns rather than a
separate measurement column for every visit.

For upload, the required columns are `USUBJID`, `PARAMCD`, `PARAM`, `AVISIT`,
`AVISITN`, `AVAL`, and `AVALU`, together with a supported treatment variable
such as `TRT01A`. `AVAL` and `AVISITN` must be numeric. The supplied file already
has these columns.

Open **Data & profile** to inspect the active data context. A successful upload
does not mean the selected records have passed all plotting checks; those checks
run after you confirm the choices.

## 2. Describe the question

On the **Ask** step, keep the supplied example question:

> Create a boxplot of ALT AVAL by visit, split by treatment. Include all visits and treatment groups, show the number of subjects below each box, and connect the medians.

Select **Continue to review choices**.

The current upload workflow records this question with the revision. It derives
the available choices from the uploaded data and uses your confirmed selections
to define the plot. It does not interpret the question through an LLM. Check each
selection even when the question describes exactly what you want.

## 3. Confirm the plot

Under **Confirm plot choices**, use these settings:

| Control | Selection |
| --- | --- |
| Parameter | `ALT` |
| Y variable | `AVAL` |
| Unit | `U/L` |
| Treatment facet | `TRT01A` |
| Included treatment levels | `Placebo` and `Active` |
| Included visits | `Baseline`, `Week 4`, and `Week 8` |
| Y scales | **Fixed** |

Select **Confirm and generate**.

The application validates the selected records against its supported BDS
plotting profile before generating the R script. Blocking diagnostics stop
generation. These checks assess the supported plotting envelope, not full ADaM
conformance. Analysis flags are not silently applied as filters; this example
uses the records selected through the controls.

Keep **Fixed** Y scales for this walkthrough. **Free** scales require an explicit
acknowledgement and create an `Experimental/Draft` revision. That path does not
enter governed subprocess execution, verification, or controlled export.

## 4. Inspect the result

On **Result**, inspect the **Plot preview**. For this example, look for ALT
measurements at the three selected visits, split into the two treatment facets.
The plot shows boxplots, visible outliers, connected medians, and distinct-subject
N values in a strip below the boxes.

Open **Executed R script** to inspect the generated code. Open
**Optional: view automated checks** to inspect the evidence and check messages.
Read any nonblocking profile warnings shown in **Run Status**.

For fixed scales, the application generates its own R script and runs it
synchronously in a clean `callr` subprocess. Passing execution-service checks
accepts the artifacts and moves the revision to `Verified`. A failed check can
leave the revision at `Draft`; read the diagnostic before continuing.

| Status | Meaning for this walkthrough |
| --- | --- |
| `Draft` | The revision has not reached successful execution verification. |
| `Verified` | The execution-service checks passed; human review remains separate. |
| `Reviewed` | Both required independent reviewer roles approved the revision's artifacts. |
| `Rejected` | A reviewer rejected the revision; it cannot be exported. |
| `Experimental/Draft` | The free-scale path is experimental and cannot be exported. |

`Verified` does not establish statistical correctness or human approval.
The UI preview is assembled in the application process; it is not the accepted
PNG produced by the subprocess. Inspect the downloaded PNG as well.

## 5. Review and export

Select **Continue to export**. Expand **Traceability, review, and export**.

For a demonstration of human review, use **Two-Person Review**:

1. Select **Statistical programmer** for both **Reviewer identity** and
   **Review role**. Inspect the result and evidence, enter a comment, and select
   **Approve** if the result is acceptable.
2. Have the second reviewer inspect the same revision. Select **Biostatistician**
   for both identity and role, then record that reviewer's decision.

Both distinct non-creator actors must approve the same artifact hashes for the
revision to become `Reviewed`. The POC identities are selectable local actors,
not authenticated user accounts. Selecting both identities yourself demonstrates
the mechanics; it is not independent human review.

Under **Controlled export**:

1. Select **Creator** as **Actor** for this local exercise.
2. Keep **Destination** as `local` when using the default runtime configuration.
3. Select **Export image and code**.
4. Confirm that the message reports **Exported** and a receipt appears.
5. Select **Download image and code ZIP**.

Export must succeed before the ZIP download can be created. The destination is
a registered workspace identifier, not a folder path. If your runtime
configuration uses another workspace identifier, use that registered destination.

The ZIP contains exactly two files:

| File | Content |
| --- | --- |
| `plot.png` | The accepted plot image from the exported revision. |
| `script.R` | The accepted R script from the exported revision. |

The export receipt stays in the internal evidence store. The ZIP does not
contain the uploaded dataset, review decisions, or a complete replay bundle.
Download only after successfully exporting the revision you intend to keep:
the ZIP packages the last successful export.

The current service permits `Draft`, `Verified`, and `Reviewed` exports when an
accepted artifact bundle and the other export checks are present. It does not
require both approvals for every export. Treat any unreviewed output as a draft;
the detached files do not carry the revision's review-status label. Do not infer
approval from a successful export.

## Check what you downloaded

Unzip the download and open `plot.png`. Confirm the parameter, unit, visits,
treatment facets, and N strip against your selections. Open `script.R` and
inspect it before running it in another environment.

The generated script has a checkout-specific fallback data path and package
installation code. Copying the script and PNG alone does not guarantee portable
reproduction. Reproduction requires the appropriate input records, dependencies,
and execution environment.

The local POC stores session evidence and workspace outputs under temporary
directories. Keep the downloaded files you need; do not assume a later session
will provide access to this revision or its internal evidence.

## If a step stops

| Symptom | Next action |
| --- | --- |
| Upload does not show **File ready** | Read the upload message; check the required columns and numeric fields. For Excel, check the first worksheet. |
| Generation reports a blocking profile diagnostic | Resolve the named issue in the input or selections before generating again. |
| The result remains `Draft` | Inspect the execution diagnostics and automated checks; do not treat the preview as verification evidence. |
| Export fails | Read the export message and check the status, actor, destination, and accepted artifact availability. |
| ZIP download reports that export is required | Complete **Export image and code** successfully before downloading. |

## Implementation references

This guide describes the inspected source behavior, not a record of a completed
browser test. The main implementation references, relative to the repository
root, are:

- `R/mod_plot_generation.R`: `mod_plot_generation_ui()` and
  `.uploaded_study_data_provider()` define the journey and upload adapter.
- `R/mod_specification.R`: `mod_specification_server()` collects plot choices.
- `R/app_server.R`: `.create_assurance_candidate()` and `.materialize_revision()`
  connect the question, selections, execution, and revision state.
- `R/mod_plot_preview.R`: `mod_plot_preview_server()` renders the preview and code.
- `R/mod_review.R`: `mod_review_server()` records review decisions.
- `R/export_service.R`: `new_export_service()` checks and publishes the export.
- `R/mod_export.R`: `mod_export_server()` and `.write_export_zip()` provide export
  controls and the two-file ZIP download.
